Toward Resolved Simulations of Burning Fronts in Thermonuclear X-ray Bursts

### 2018-10-02 astronum_2018/astronum-xrb.tex
Add paragraph about VODE timestepping and the restart experiment.

\newcommand{\maestro}{{\sffamily Maestro}}
\newcommand{\castro}{{\sffamily Castro}}
\newcommand{\starkiller}{{\sffamily StarKiller}}
\newcommand{\starlord}{{\sffamily StarLord}}
\newcommand{\nyx}{{\sffamily Nyx}}
\newcommand{\amrex}{{\sffamily AMReX}}

@@

As an example of the improvements in coupling between hydro
and \MarginPar{Don can add some comments about startup costs?}
reactions, Figure~\ref{fig:sdc} shows the mass fraction of helium
behind a detonation over the course of 2 timesteps.  For this test,
a 19 isotope reaction network was used.  The points
represent individual calls to the righthand side function of our
reaction network, and the density of
points indicates how hard the integrator is working.  We use the same
tolerances for both the Strang-split method and the SDC method.  For
the SDC method, we are using a second-order accurate method based on
\cite{SDC-old}.  We predict a time-centered advection term,
$\Adv{\Uc}^{n+1/2}_i$ using standard unsplit Godunov methods, but
explicitly include a reactive source term in the tracing of the
interface states, ${\bf R}(\Uc_i)$.  We can then solve the reactive
system, using this advective term as a piecewise-constant-in-time
source
\begin{equation}
\label{eq:syst}
\frac{d{\Uc_i}}{dt} = -\Adv{\Uc}^{n+1/2}_i + {\bf R}(\Uc_i)
\end{equation}
This system can be integrated with standard ODE methods.  To achieve
second-order accuracy, we need to iterate, using the updated $\Uc$ to
create a reactive source included in the predictor for the advective
update, and then reintegrate the system.  The result of this iteration
is that the advection sees the effects of reactions over the timestep
and the reactions see the effects of advection.  For the zone tracked
in Figure~\ref{fig:sdc}, each SDC iteration called the righthand side
function of the network 2 times less than the Strang case.  This
savings is dependent on the network and thermodynamic state, but our
tests thus far show that the SDC methods need fewer network calls to
integrate a timestep.  We also clearly see how the Strang system
evolves in a discontinuous fashion.  We are continuing to develop this
new integration strategy, with full fourth-order in space and time
reacting hydrodynamics methods to follow shortly.  The XRB simulations
are our target application.

>>>>

\newcommand{\maestro}{{\sffamily Maestro}}
\newcommand{\castro}{{\sffamily Castro}}
\newcommand{\starkiller}{{\sffamily StarKiller}}
\newcommand{\starlord}{{\sffamily StarLord}}
\newcommand{\nyx}{{\sffamily Nyx}}
\newcommand{\amrex}{{\sffamily AMReX}}
\newcommand{\vode}{{\sffamily VODE}}

@@

As an example of the improvements in coupling between hydro
and reactions, Figure~\ref{fig:sdc} shows the mass fraction of helium
behind a detonation over the course of 2 timesteps.  For this test,
a 19 isotope reaction network was used.  The points
represent individual calls to the righthand side function of our
reaction network, and the density of
points indicates how hard the integrator is working.  We use the same
tolerances for both the Strang-split method and the SDC method.  For
the SDC method, we are using a second-order accurate method based on
\cite{SDC-old}.  We predict a time-centered advection term,
$\Adv{\Uc}^{n+1/2}_i$ using standard unsplit Godunov methods, but
explicitly include a reactive source term in the tracing of the
interface states, ${\bf R}(\Uc_i)$.  We can then solve the reactive
system, using this advective term as a piecewise-constant-in-time
source
\begin{equation}
\label{eq:syst}
\frac{d{\Uc_i}}{dt} = -\Adv{\Uc}^{n+1/2}_i + {\bf R}(\Uc_i)
\end{equation}
This system can be integrated with standard ODE methods.  To achieve
second-order accuracy, we need to iterate, using the updated $\Uc$ to
create a reactive source included in the predictor for the advective
update, and then reintegrate the system.  The result of this iteration
is that the advection sees the effects of reactions over the timestep
and the reactions see the effects of advection.  For the zone tracked
in Figure~\ref{fig:sdc}, each SDC iteration called the righthand side
function of the network 2 times less than the Strang case.  This
savings is dependent on the network and thermodynamic state, but our
tests thus far show that the SDC methods need fewer network calls to
integrate a timestep.  We also clearly see how the Strang system
evolves in a discontinuous fashion.  We are continuing to develop this
new integration strategy, with full fourth-order in space and time
reacting hydrodynamics methods to follow shortly.  The XRB simulations
are our target application.

For the ODE integrations shown we use the variable-order,
variable-step integrator \vode\ \cite{vode} with its implementation of
the backward-differentiation formulas. The integration order
corresponds to the order of the interpolating polynomials used to
predict the solution in each integrator step. We note that for both
Strang splitting and SDC, \vode\ must start the integration at first
order and correspondingly small time steps since no higher-order
information about the solution is available. That is, our reaction
network right-hand sides provide the first time derivative of the
solution but no higher derivatives. Based on testing with a 13-isotope
approximate alpha-chain network, \vode\ takes many small timesteps
(e.g. $\approx 200$) before it can stably increase to a high order
representation of the solution, but after it does so, it can often
complete the integration in only a handful of steps (e.g.
$\approx 5-6$). Observing that during a \castro\ timestep, the first
Strang step is a direct continuation of the integration in the second
Strang step of the previous \castro\ timestep, we explored saving the
\vode\ state following the second Strang step and restarting \vode\
during the first Strang step. This approach allowed us to start the
first Strang step with a high order representation of the solution and
complete it in very few \vode\ timesteps (e.g. $<10$ steps), speeding
up the \castro\ timestep advance by 50\% for our detonation test
problem using the 13-isotope network. However, this came at the memory
cost of storing $7 N_{eq} + 90$ scalar variables on our grid, where
for our networks $N_{eq} = N_{spec} + 2$ (in addition to the number of
species $N_{spec}$ we also integrate temperature and specific
energy). For a large-scale science problem with a large network, this
could quickly become prohibitive, so we are exploring ways to mitigate
either the memory requirements for restarting the integration or the
small timestep start up costs for the integration method.
